Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-06 paper/tidy-missing-data-paper.Rmd
removed many unnecessary instances of "that" and general grammar clean up.

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
  # For running headers, if needed
  short:     "\\pkg{naniar}: Explore missing data"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with a goal to integrating missing value handling as an integral part of data analysis workflows. New data structures are defined along with new functions (verbs) to perform common operations. Together these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

Tidy data [@Wickham2014] is a relatively new principle and suite of tools that
facilitate the process of converting raw data into a clean, analysis-ready
format that works efficiently in a data analysis pipeline. In tidy data, missing
values can be handled implicitly or explicitly, at the analyst's discretion.
However, when making a plot, missing values are simply dropped, albeit with a
warning (see Figure \@ref(fig:warning)). Most other analysis software will
simply drop cases with missings without warning.

@@

The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
@Schafer2002, @VanBuuren2012) does not address how to handle missing data.
Instead the focus is on ensuring valid statistical inference is made from
incomplete data. They approach this chiefly through probabilistic modelling, and
assume the mechanism of missing data is known to the analyst.

However, something has to be known about the missing value structure to produce
a complete dataset for analysis. Whether by case or variable deletion, or with
imputation, it is ideal to know what the mechanism is that generates missing 
values. To understand their structure, the data must be explored, which means generating visualizations, summaries, and performing exploratory modelling. This process is inherently difficult, as it is iterative, and
involves many dead ends, and there might not be a clear answer even after much
exploratory data analysis.

Although
decision tree models or latent group analysis can reveal structures or patterns
of missingness [@Tierney2015; @Barnett2017], these are not definitive and
usually require discussion with data curators to validate.
While there might be other times where the data are
clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet for missing data that will reveal
it's mechanism to the analyst. Missing data must be explored.

While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs that make them challenging to integrate into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on the concepts that make
this process more efficient, and integrate well into analysis pipelines.

@@

The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analysts capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

The paper is organized in the following way to discuss the problem of missing data. The next section (\@ref(tidy-data-concepts)) is an introduction to tidy data principles and tools. Missing value imputation (\@ref(missing-data-rep-dep) and \@ref(imputation)) and exploration literature (\@ref(exploration)) are then discussed. Section \@ref(extensions) discusses extensions to the tidy methodology that facilitates operations for exploring, visualizing, and imputing missing data. The application of these new methods is illustrated using a case study in section \@ref(case-study). Finally, section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors.

@@

Data manipulation is made input and output tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. **Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming** changes existing variables or adding new ones; **aggregating** creates a single value from many values, say for example by computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. The \pkg{recipes} package is a tidy tool under development to help make modelling input and output tidy [@recipes].

@@

Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@tidyverse]. These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to a growth of tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr}, dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to the growth of new packages for other fields that follow similar design principles and create fluid workflows for new domains. For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].

@@

Past graphics research on missing data has not yet been framed in a tidy way.
The graphics literature has explored methods for missing value exploration and imputation. Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows. By laying down these concepts, it allows for the focus to be more general than just software, but rather, a framework for tidy tools to explore missing data that can be built upon, extended, and implemented in other languages.

@@

**MCAR** is where values being missing have no association with observed or unobserved data, that is $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness. Although this is a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.

@@

Finally, **MNAR** refers to the missingness being related to values observed, and un unobserved; if this does not simplify  $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself. This presents a challenge in analysis, as it is difficult to verify this state, and also implies bias in analysis due to the unobserved phenomena.

@@

\pkg{VIM} [@VIM] efficiently implements well-used imputation methods k-nearest neighbor, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allow the package to deal semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration). The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.

A key feature of the design of \pkg{simputation} is that it returns a regular dataframe. This is different from other imputation packages, which return custom classes of data or additional objects. These create extra work when preparing for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost identifying imputed values.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts that do cover exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

VIM (Visualizing and Imputing Missing Data) provides visualization methods that identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

@@

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(vis-summaries) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems that arise
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.

@@

To explore missing data there needs to be a good representation of missing
values that integrate into analysis pipelines. The R matrix, where 0 and 1
indicate not missing and missing provides a template, as it is very common in
missing data literature.  This matrix has been used to explore missing values in
the interactive graphics library XGobi, where it was called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness, where 1s represent missing values, and 0
represents complete values. The values 0 and 1 (or equally, FALSE and TRUE) can
be confusing, however, as it is not immediately clear what they represent - does
0 indicate a the absence of observation, or the presence of a missing value? The
shadow matrix was specially linked to the data, but could not be referred to
directly, instead they were linked with special window views. This meant that
the data structure was abstracted from the user to facilitate data exploration.
This made sense at the time with the state of the art for graphicaly analysis,
but that data structure needs to be clearly presented to the user so that they
can more broadly use it in other data analyses.

@@

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction that `var_NA` is a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

@@

knitr::kable(fig_shadow_code,
             caption = "Example data of temperature, and its observed value, shadow representation, and underlying value. The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data that is otherwise difficult to record.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats that have quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses visualizations that are easy to remember and compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.

@@

All missings in a dataset can be displayed using a heatmap style visualization. This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is that `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data that is similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)). Figure \@ref(fig:airquality-upset) shows that only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows that Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

@@

airquality %>%
  bind_shadow() %>%
  impute_below_all() %>%
  ggplot(aes(x = Ozone,
             fill = Ozone_NA)) + 
  geom_histogram() + 
  scale_fill_brewer(palette = "Dark2") + 
  theme_minimal()
```

@@

airquality %>%
  bind_shadow() %>%
ggplot(aes(x = Temp)) + 
  geom_histogram(na.rm = TRUE) + 
  facet_wrap(~Ozone_NA) + 
  labs(tag = "A")

airquality %>%
  bind_shadow() %>%
ggplot(aes(x = Temp,
           colour = Ozone_NA)) + 
  geom_density(na.rm = TRUE) +
  scale_colour_brewer(palette = "Dark2")  + 
  labs(tag = "B")

@@

```{r geom-miss-point-impute-shift-long, include = FALSE, out.width = "70%"}
airquality %>%
  bind_shadow() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  ggplot(aes(x = Ozone,
             y = Solar.R,
             colour = any_missing)) + 
  geom_point() + 
  scale_colour_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom")
```

@@

Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values. Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows that humidity is missing at low air and sea temperatures, and that humidity is missing in one year, and one location.

@@

dat_paral <-  oceanbuoys %>%
  mutate(ID = 1:n(),
         ID = as.factor(ID)) %>%
  bind_shadow() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  mutate_if(is.numeric, range_01) %>%
  gather(key, value, -c(9:19)) %>%
  select(ID, key, value, everything())

@@

The overall number of missing values in a dataset is shown with `n_miss`, and the proportion or percent missings, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These be extended to summarize the number/amount of variables and cases that contain a missing values by appending `_case` or `_var` to any of these summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.

@@

There are many common values that mean "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count). All occurrences of -99 can be searched for using the code:

@@

One drawback to imputation functions in this form is that the location of imputed values is not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).

>>>>

---
documentclass: jss
author:
  - name: Nicholas Tierney
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics
      E774, Menzies Building
      Monash University, Clayton
    email: \email{nicholas.tierney@gmail.com}
    url: http://www.njtierney.com
  - name: Dianne Cook
    affiliation: Monsh University
title:
  formatted: "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package \\pkg{naniar}"
  # If you use tex in the formatted title, also supply version without
  plain:     "Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations with the Package naniar"
  # For running headers, if needed
  short:     "\\pkg{naniar}: Explore missing data"
abstract: >
  Despite the large body of research on missing value distributions and imputation, there is comparatively little literature with a focus on how to make it easy to handle, explore, and impute missing values in data. This paper addresses this gap. The new methodology builds upon tidy data principles, with the goal of integrating missing value handling as an integral part of data analysis workflows. We define both new data structures and new verbs (functions) to perform common operations. Together, these provide a cohesive framework for handling, exploring, and imputing missing values. These methods have been made available in the R package `naniar`.
keywords:
  # at least one keyword must be supplied
  formatted: [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, "\\proglang{R}"]
  plain:     [workflow, statistical computing, data science, data visualization, tidyverse, data pipeline, R]
preamble: >
  \usepackage{amsmath}
  \usepackage[english]{babel}
output:
  bookdown::pdf_book:
    base_format: rticles::jss_article
    includes:
      in_header: header.tex
    keep_tex: yes
    citation_package: 'natbib'
bibliography: tidy-missing-data-paper.bib
# csl: jcgs.csl
---

@@

As data science has become a more solid field, theories and principles have developed to describe best practices. One such idea is 'tidy data,' a set of guidelines for converting raw data into a clean, analysis-ready format that works efficiently in a data analysis pipeline, while another is the idea of the Grammar of Graphics, which describes how to map data values into a visualization [@Wickham2014, Wilkinson2012]. These principles have guided analysts in their work for several years now. However, while these principles outline how to work with data in a general way, they do not address the problem of missing data. In particular, there is little to no guidance for working with missing values in the field of visualization. Most analysis software simply drops missing values when making a plot, although some packages provide a warning (see Figure \@ref(fig:warning)).

<!-- Tidy data [@Wickham2014] is a relatively new principle and suite of tools -->
<!-- facilitating the process of converting raw data into a clean, analysis-ready -->
<!-- format that works efficiently in a data analysis pipeline. In tidy data, missing -->
<!-- values can be handled implicitly or explicitly, at the analyst's discretion. -->
<!-- However, when making a plot, missing values are simply dropped, albeit with a -->
<!-- warning (see Figure \@ref(fig:warning)). Most other analysis software will -->
<!-- simply drop cases with missings without warning. -->

@@

The imputation literature (e.g. @Little1988, @Rubin1976, @Simon1986,
@Schafer2002, @VanBuuren2012) does not address how to explore, understand, and handle
missing data structures and mechanisms.
Instead the focus is on ensuring valid statistical inference is made from
incomplete data. They approach this chiefly through probabilistic modelling, and
assume the mechanism of missing data is known to the analyst.

However, something has to be known about the missing value structure to produce
a complete dataset for analysis. Whether by case or variable deletion, or with
imputation, it is ideal to identify the mechanism generating missing 
values. To understand their structure, the data must be explored, which means generating visualizations, summaries, and performing exploratory modelling. This process is inherently difficult, as it is iterative, and
involves many dead ends, and there might not be a clear answer even after much
exploratory data analysis.

Although
decision tree models or latent group analysis can reveal structures or patterns
of missingness [@Tierney2015; @Barnett2017], these are not definitive and
usually require discussion with data curators to validate.
While there might be other times where the data are
clearly missing completely at random (or missing at random), determining this is not always straightforward. There is no silver bullet to reveal missing data
mechanisms to the analyst. Missing data must be explored.

While there have been many software tools for exploring missing data (VIM, mi, mice, MissingDataGUI, MANET, ggobi, norm), these tools generate outputs challenging to integrate into analysis pipelines. Despite it being crucial to explore and understand missing data, there is very little research on how to make this process more efficient and integrate it into the analysis pipeline.

@@

The graphics literature [@Swayne1998; @Unwin1996; @VIM; @Cheng2015] provides several solutions for exploring missings visually. These approaches incorporate missings into the plot in some way. For example, imputing values 10% below the range to display all values in a scatter plot. These methods require the data to be linked to the missing or imputed values; this link is typically hidden from the user. The approaches from the graphics literature have not been translated and integrated into tidy data. This limits the analyst's capacity to efficiently explore missing values. Extending tidy data and tidy tools to account for missing values would create more efficient and robust workflows and analysis.

The paper is organized in the following way to discuss the problem of missing data. The next section (\@ref(tidy-data-concepts)) is an introduction to tidy data principles and tools. Missing value imputation (\@ref(missing-data-rep-dep) and \@ref(imputation)) and exploration literature (\@ref(exploration)) are then discussed. Section \@ref(extensions) discusses tidy methodology extensions, which facilitate operations for exploring, visualizing, and imputing missing data. The application of these new methods is illustrated using a case study in section \@ref(case-study). Finally, section \@ref(discussion) discusses strengths, limitations, and future directions.

@@

Tidy data is easier to work with and analyze because the variables are in the same format as they would be put into modelling software. This helps the analyst works swiftly and clearly, closing up opportunities for errors. Tidy data principles are general, but are most robustly implemented in the R programming language, so the 
paper focusses on R.

@@

Data manipulation is made input- and output-tidy with R packages \pkg{dplyr} and \pkg{tidyr} [@dplyr; @tidyr]. These provide the five "verbs" of data manipulation: data reshaping, sorting, filtering, transforming, and aggregating. **Data reshaping** goes from long to wide formats; **sorting** arranges rows in a specific order; **filtering** removes rows based on a condition; **transforming** changes existing variables or adding new ones; **aggregating** creates a single value from many values, say for example by computing the minimum, maximum, and mean.

@@

Modelling tools work well with tidy data, as they have a clear mapping from variables in the data to the formula for a model. For example in R, y regressed on x and z is: \code{lm(y ~ x + z)}. Modelling tools are input tidy, but their output is always messy - it is not in the right format for subsequent steps in analysis. For example, estimated coefficients, predictions, and residuals from one model cannot be easily combined with the output of another model. This output problem has been partially addressed with the \pkg{broom} package, which tidies up model outputs for data analysis, and the developing \pkg{recipes} package, which helps make modelling input and output tidy [@recipes].

@@

Defining tidy data and tidy tools has resulted in the growth of a set of packages known collectively as the "tidyverse" [@tidyverse]. These are designed to share similar principles in their design and behavior, and cover the breadth of an analysis - going from importing, tidying, transforming, visualizing, modelling, to communicating [@r4ds; @tidyverse; @Tidyverse-Manifesto]. This has led to more tools for specific parts of analysis - from reading in data with \pkg{readr}, \pkg{readxl}, and \pkg{haven}, to handling character strings with \pkg{stringr} and dates with \pkg{lubridate}, and performing functional programming with \pkg{purrr} [@readr; @readxl; @haven; @stringr; @lubridate; @purrr]. It has also led to a burgeoning of new packages for other fields following similar design principles and create fluid workflows for new domains. For example, the \pkg{tidytext} package for text analysis, the \pkg{tsibble} package for time series data, and \pkg{tidycensus} for working with US census and boundary data [@tidytext; @wang2020tsibble; @tidycensus].

@@

Graphics research on missing data has not yet been framed in a tidy way.
The graphics literature has explored methods for missing value exploration and imputation. Translating and expanding these to fit within tidy data and tidy tools would create more efficient workflows. By laying down these concepts, it allows the focus to be more general than just software, but rather, an extensible framework for tidy tools to explore missing data.

@@

**MCAR** is where values being missing have no association with observed or unobserved data, that is, $Pr(R = 1 | Y_{obs}, Y_{miss}) = Pr(R = 1 | \psi)$. Essentially, the probability of an observation being missing is unrelated to anything else, only the parameters $\psi$, which is the overall probability of missingness. Although this is a convenient scenario, it is not actually possible to confirm, or clearly distinguish from MAR, as it relies on statements on data unobserved.

@@

Finally, **MNAR** refers to the missingness being related to values observed, and un unobserved; if this does not simplify  $Pr(R = 1 | Y_{obs}, Y_{miss}, \psi)$. This assumes that conditioning on all available observed data, the data goes missing due to some phenomena unobserved, including the structure of the missing data itself.  This presents a challenge in analysis, as it is difficult to verify, and implies bias in analysis due to the unobserved phenomena.

@@

\pkg{VIM} [@VIM] efficiently implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allows the package to deal with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration). The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.

A key feature of the design of \pkg{simputation} is that it returns a regular dataframe. This is different from other imputation packages, which return custom classes of data or additional objects. These create extra work when preparing for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost of identifying imputed values.

@@

The primary focus of most missing data packages is making inferences, and exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts covering exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

@@

VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

@@

Applying tidyverse principles to the domain of missing data clarifies and
facilitates missing data exploration, visualization, and imputation. This section
discusses how these principles are applied, and proceeds in three parts. Section
\@ref(data-structure) discusses _nabular_ data, a tidy data structure for
missing values that tracks missings throughout the analysis and allows for
special missing values. Section \@ref(vis-summaries) covers visualizations to
understand missingness. Section \@ref(num-sum) discusses data summaries, and
Section \@ref(verbs) covers the common "verbs" used to solve problems
when working with missing data. The design and naming of these features is an
important component that helps the analyst compose them efficiently in data pipelines, and is discussed throughout.

@@

To explore missing data there needs to be a robust representation of missing
values that integrate into analysis pipelines. The R matrix, where 0 and 1
indicate not missing and missing provides a template, as it is very common in
missing data literature.  This matrix has been used to explore missing values in
the interactive graphics library XGobi, where it was called a "missing value
shadow", or "shadow matrix", defined as a copy of the original data with
indicator values of missingness, where 1s represent missing values, and 0
represents complete values. The values 0 and 1 can
be confusing representations of missing values - does
0 indicate a the absence of observation, or the presence of a missing value? The
shadow matrix was specially linked to the data, but could not be referred to
directly, and was instead linked with special window views. This means
the data structure was abstracted from the user to facilitate data exploration.
This made sense at the time with the state of the art for graphical analysis,
but the data structure needs to be clearly presented to the user so they
can use it more flexibly in other data analyses.

@@

A consistent short suffix "`_NA`" for each variable keeps names coordinated throughout analysis (See Figure \@ref(fig:shadow-matrix-progress) part 4). The suffix is short and easy to remember in analysis or visualization. It also makes a clear distinction with `var_NA` being a random variable of the missingness of a variable, `var`. This subtle change is important as it shifts the focus from the value of a variable to its missingness state, making intent clear when performing analysis.

@@

knitr::kable(fig_shadow_code,
             caption = "Example data of temperature, and its observed value, shadow representation, and underlying value. The temperature value of -99 is represented in the shadow column as 'NA\\_instr', which in turn has the underlying numeric value of 2. This captures additional information about the data otherwise difficult to record.")
             # booktabs = TRUE) 
  # kable_styling(latex_options = c("hold_position"))

@@

_Nabular_ data binds the shadow matrix column-wise to the original data, and is so named as a portmanteau of `NA` and `tabular`. _Nabular_ data keeps corresponding rows together, removing the possibility of mismatching records, explicitly linking missing values to the data. _Nabular_ data facilitates visualization and summaries by allowing the user to reference the missingness of a variable, `var`, as `var_NA`. _Nabular_ data is a snapshot of the missingness of the data. This means when _nabular_ data is imputed, those imputed values can easily be identified in analysis by referring to the appropriate, coordinated names. _Nabular_ data is created using the command `bind_shadow()` or `nabular()` (see Figure \@ref(fig:nabularfig)). _Nabular_ data is not unlike classical data formats with quality or flag columns associated with each measured variable, e.g. Scripps CO2 data [@Keeling2005-scripps], GHCN data [@Durre2008-ghcn].

@@

Visualizing missingness in a dataset is essential to understand its structure. This section discusses easy to remember visualizations that compose with other functions in a pipeline. Four different areas of visualization of missing data are discussed: overviews (Section \@ref(overview-plots)); univariate (Section \@ref(univariate-plots)), bivariate (Section \@ref(bivariate-plots)), and multivariate (Section \@ref(multivariate-plots)). All plots are created using \pkg{ggplot2}, giving users clearer control over the plot appearance, and customizability.

@@

All missings in a dataset can be displayed using a heatmap style visualization. This is achieved using `vis_miss()` from the \pkg{visdat} package [@visdat], which also provides summaries of missingness overall in the legend, and for each column (see Figure \@ref(fig:vis-miss)). The user can also apply clustering to the rows, and arrange columns by missingness (Figure \@ref(fig:vis-miss)B). Similar visualizations are available in other missing data packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns in the visualization and are named at the top, and each row is an observation. Using \pkg{ggplot2} means the plot can be customized and combined with other ggplot graphics.

@@

The number of times certain variables go missing together can be visualized using an "upset plot" [@Conway2017], a visualization technique for showing the size and features of sets in data similar to a venn diagram, but scales better with more variables (Figure \@ref(fig:airquality-upset)). Figure \@ref(fig:airquality-upset) shows only Ozone and Solar.R have missing values. The dots in the bottom right show the combinations of missingness, with the bottom left showing the missingness in each variable, and the top showing the missingness for the combinations. The bottom left shows Ozone has the most missing values, and the top shows 2 cases where both Solar.R and Ozone have missing values together.

@@

airquality %>%
  nabular() %>%
  impute_below_all() %>%
  ggplot(aes(x = Ozone,
             fill = Ozone_NA)) + 
  geom_histogram() + 
  scale_fill_brewer(palette = "Dark2") + 
  theme_minimal()
```

@@

airquality %>%
  nabular() %>%
ggplot(aes(x = Temp)) + 
  geom_histogram(na.rm = TRUE) + 
  facet_wrap(~Ozone_NA) + 
  labs(tag = "A")

airquality %>%
  nabular() %>%
ggplot(aes(x = Temp,
           colour = Ozone_NA)) + 
  geom_density(na.rm = TRUE) +
  scale_colour_brewer(palette = "Dark2")  + 
  labs(tag = "B")

@@

```{r geom-miss-point-impute-shift-long, include = FALSE, out.width = "70%"}
airquality %>%
  nabular() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  ggplot(aes(x = Ozone,
             y = Solar.R,
             colour = any_missing)) + 
  geom_point() + 
  scale_colour_brewer(palette = "Dark2") + 
  theme(legend.position = "bottom")
```

@@

Figure \@ref(fig:parallel-cord-plot) shows a parallel coordinate plot of the `oceanbuoys` data, with missing values imputed to be 10% below the range, and values colored according to whether humidity contained missing values. Although a 10% below imputation is not the most ideal way to display missing values in a parallel coordinate plot, it shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

@@

dat_paral <-  oceanbuoys %>%
  mutate(ID = 1:n(),
         ID = as.factor(ID)) %>%
  nabular() %>%
  impute_below_all() %>%
  add_label_shadow() %>%
  mutate_if(is.numeric, range_01) %>%
  gather(key, value, -c(9:19)) %>%
  select(ID, key, value, everything())

@@

The overall number of missing values in a dataset is shown with `n_miss`, and the proportion or percent missings, `prop_miss` and `pct_miss`. Complementing these are functions to reveal the number of complete values: `n_complete`, `prop_complete`, and `pct_complete`. These be extended to summarize the number/amount of variables and cases containing a missing values by appending `_case` or `_var` to any of these summaries. Table \@ref(tab:n-prop-pct-miss-complete) shows the names and output of these functions on the dataset airquality.

@@

There are many common values for "missing". For example, "N/A" "Not Available", "MISSING", or white space " ". Before replacing these values with missing values (`NA`), it is useful to first _scan_ the data to get a sense of the magnitude of the problem. The  `miss_scan_count()` function finds where specified values occur in data. For example, for the example data, `dat_ms`  Table \@ref(tab:setup-miss-scan-count). All occurrences of -99 can be searched for using the code:

@@

One drawback to imputation functions in this form is imputed value locations are not tracked. This issue is resolved with \code{nabular} data covered in Section \@ref(verbs-track).
